Bring More Buses: Transit Delays in the Toronto Transit Commission

### 2024-01-26 outputs/paper/paper.qmd
paper finished!

The "TTC", or Toronto Transit Commission, has been Toronto's transit system since 1921, serving the city population with mainly subways, streetcars, buses, along with smaller-scale transport offerings. The openly available data surrounding the delays of these various modes of transport offers opportunity to investigate the root causes of the longer-lasting delays. Delays are likely the best way to assess a transit agency's performance from both the rider and management perspective, since riders' experience is typically determined through how low waiting and travel times are, while performance of staff is assessed "in terms of 
on-time performance standards (or adherence to schedules)" [@diab2015bus]. Both of these metrics are highly correlated with delay time. Therefore, understanding what caused a delay, when it was caused and where it was caused could give the TTC more insight into how to deliver their services to the public more efficiently and effectively.

@@

```{r fig-reasons-delay}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Distribution of Bus and Streetcar Delay Reasons 
#| fig-align: center

@@

Next, the variation of delay incidents by day of the week was analyzed. Figure @fig-dow-1 shows the number of delay incidents for each day.

```{r fig-dow-1}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Number of Delay Incidents, aggregated by Day of Week
#| fig-align: center
#Day of week with most incidence of delay

@@

```{r fig-dow-2}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Average Delay Time, aggregated by Day of Week
#| fig-align: center

@@

#plot
avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = Day, y = `avg_delay`)) +
  geom_point() +
  geom_line(group = 1) + 
  labs(title="Average Delay Time by Day of Week", x="Day of Week", y="Average Delay (mins)") +
  theme_minimal() +
  ylim(0, max(avg_time_delayed_vehicle$avg_delay) + 1) + 
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5))
```

Next, we looked at average delay duration by day. Figure @fig-dow-2 displays the average delay time in minutes for each day. The average remains relatively consistent across the week, hovering around 15 minutes per delay incident.

@@

Next, we graph the data to check for noticeable differences across different genres of transportation. We see a noticeable difference in delays by vehicle in Figures @fig-vehicle-1 and @fig-vehicle-2

```{r fig-vehicle-1}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Number of Incidents for each mode of transportation
#| fig-align: center

@@

@@

@@

Figure @fig-vehicle-1 shows that in 2023, buses had 56207 recorded delay incidents in 2023. This is more than twice as many as subways and close to three times more than streetcars.

```{r fig-vehicle-2}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Average Delay Time for each Mode of Transportation
#| fig-align: center

avg_time_delayed_vehicle <- cleaned_transit_data |> group_by(vehicle) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")

avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = vehicle, y = avg_delay)) +
  geom_col() + 
  labs(title="Average Delay by Vehicle", x="Vehicle Type", y="Average Delay (mins)") +
  theme_minimal() +
  theme(plot.title.position = "plot",
      plot.title = element_text(hjust = 0.5))
```

In Figure @fig-vehicle-2, however, we do not see as large a difference in average delay times across different genres of transportation, however. Buses had the longest average delay times, at around 20 minutes, but subway had only 3 minutes of delay time on average, making for a fairly sizeable gap delay times. Interestingly, although there was less incidence of streetcar delay than subway delay, streetcar routes tended to be delayed longer on average throughout 2023.

@@

To investigate how TTC delays change throughout the year, Figures @fig-monthly-1 and @fig-monthly-2 were constructed. The graphs show monthly changes in delay count and average delay time over 2023. Buses, streetcars, and subways are represented by blue, green, and red lines, respectively.

```{r fig-monthly-1}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Delay Incidence by Month
#| fig-align: center
#Line chart of how avg. delay time changes over the months

@@

In @fig-monthly-1 we see how the count of delays changes throughout 2023 for each mode of transportation. Buses have the highest incidence of delays throughout year with the count generally trending upward as the year progresses, peaking in November with 5322 incidents. Incidence of subway and street car delays are significantly lower, but also trend upwards as the year progresses. Overall, this figure presents similar findings to Figure @fig-vehicle-1 in section 2.3 - buses tend to have more incidences of delay than both streetcars and subways overall.

```{r fig-monthly-2}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Average Delay Time, by Month
#Line chart of how avg. delay time changes over the months

@@

Finally, the delays on particular subway lines and bus/streetcar routes were examined. Given the large number of different transit routes offered by the TTC, this analysis focuses on the 10 routes with the highest average delay times and highest incidence of delay. These findings are illustrated in Figure @fig-highest-1 and Figure @fig-highest-2. A short guide to the route number nomenclature of the TTC is provided in the Appendix for the reader's convenience.

```{r fig-highest-1}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Routes/Lines with Highest Delay Incidence
#| fig-align: center

@@

```{r fig-highest-2}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Routes/Lines with Highest Average Delay Time
#| fig-align: center

>>>>

The TTC, or Toronto Transit Commission, has been Toronto's transit system since 1921, serving the city population with mainly subways, streetcars, buses, along with smaller-scale transport offerings. The openly available data surrounding the delays of these various modes of transport offers opportunity to investigate the root causes of the longer-lasting delays. Delays are likely the best way to assess a transit agency's performance from both the rider and management perspective, since riders' experience is typically determined through how low waiting and travel times are, while performance of staff is assessed "in terms of 
on-time performance standards (or adherence to schedules)" [@diab2015bus]. Both of these metrics are highly correlated with delay time. Therefore, understanding what caused a delay, when it was caused and where it was caused could give the TTC more insight into how to deliver their services to the public more efficiently and effectively.

@@

```{r fig-reasons-delay}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Distribution of Bus and Streetcar Delay Reasons 
#| fig-align: center

@@

Next, the variation of delay incidents by day of the week was analyzed. @fig-dow-1-1 shows the number of delay incidents for each day.

```{r fig-dow-1}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Delays, aggregated by Day of Week
#| fig-subcap: ["Incidences","Average Delay Time (mins)"]
#| layout-ncol: 2
#| fig-align: center
#Day of week with most incidence of delay

@@

#Average Time Delay

@@

#plot
avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = Day, y = `avg_delay`)) +
  geom_point() +
  geom_line(group = 1) + 
  labs(title="Average Delay Time by Day of Week", x="Day of Week", y="Average Delay (mins)") +
  theme_minimal() +
  ylim(0, max(avg_time_delayed_vehicle$avg_delay) + 1) + 
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5))

Next, we looked at average delay duration by day. @fig-dow-1-2 displays the average delay time in minutes for each day. The average remains relatively consistent across the week, hovering around 15 minutes per delay incident.

@@

Next, we graph the data to check for noticeable differences across different genres of transportation. We see a noticeable difference in delays by vehicle in @fig-vehicle-1

```{r fig-vehicle-1}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Delays for each mode of transportation
#| fig-align: center
#| layout-ncol: 2
#| fig-subcap: ["Number of Incidences","Average Delay Time"]

@@

# kable(num_incidents_vehicle, format = "markdown") |> kable_styling(latex_options = c("scale_down"))

@@

avg_time_delayed_vehicle <- cleaned_transit_data |> 
                            group_by(vehicle) |> 
                            summarise(avg_delay = round(mean(`Min Delay`, na.rm = TRUE), digits = 1), .groups = "drop")

# kable(avg_time_delayed_vehicle, format = "markdown") |> kable_styling(latex_options = c("scale_down"))

avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = vehicle, y = avg_delay)) +
  geom_col() +
  labs(title="Average Delay by Vehicle", x="Vehicle Type", y="Average Delay (mins)") +
  theme_minimal() +
  theme(plot.title.position = "plot",
      plot.title = element_text(hjust = 0.5))

@@

@fig-vehicle-1-1 shows that in 2023, buses had 56207 recorded delay incidents in 2023. This is more than twice as many as subways and close to three times more than streetcars.

In @fig-vehicle-1-2, however, we do not see as large a difference in average delay times across different genres of transportation, however. Buses had the longest average delay times, at around 20 minutes, but subway had only 3 minutes of delay time on average, making for a fairly sizeable gap delay times. Interestingly, although there was less incidence of streetcar delay than subway delay, streetcar routes tended to be delayed longer on average throughout 2023.

@@

To investigate how TTC delays change throughout the year, @fig-monthly-1 and @fig-monthly-2 were constructed. The graphs show monthly changes in delay count and average delay time over 2023. Buses, streetcars, and subways are represented by blue, green, and red lines, respectively.

```{r fig-monthly-1}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Delay Incidence by Month
#| fig-align: center
#Line chart of how avg. delay time changes over the months

@@

In @fig-monthly-1 we see how the count of delays changes throughout 2023 for each mode of transportation. Buses have the highest incidence of delays throughout year with the count generally trending upward as the year progresses, peaking in November with 5322 incidents. Incidence of subway and street car delays are significantly lower, but also trend upwards as the year progresses. Overall, this figure presents similar findings to @fig-vehicle-1 in section 2.3 - buses tend to have more incidences of delay than both streetcars and subways overall.

```{r fig-monthly-2}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Average Delay Time, by Month
#Line chart of how avg. delay time changes over the months

@@

Finally, the delays on particular subway lines and bus/streetcar routes were examined. Given the large number of different transit routes offered by the TTC, this analysis focuses on the 10 routes with the highest average delay times and highest incidence of delay. These findings are illustrated in @fig-highest-1 and @fig-highest-2. A short guide to the route number nomenclature of the TTC is provided in the Appendix for the reader's convenience.

```{r fig-highest-1}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Routes/Lines with Highest Delay Incidence
#| fig-align: center

@@

```{r fig-highest-2}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Routes/Lines with Highest Average Delay Time
#| fig-align: center
